---
title: "HIV - Survival Analysis"
author: "Tharina Messeroux"
date: "2024-05-02"
output: 
  html_document:
    toc: true
    toc_float: true
---

## HIV Project 

## Loading Libraries
```{r libraries}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(tidyverse)
library(readxl)
library(knitr)
library(ggplot2)
library(gtsummary)
library(dplyr)
library(survival)
library(ggsurvfit)
library(survMisc)
```


## R Version Used 
```{r version used}
#Finding R version at use for report
R.version.string
# [1] "R version 4.3.1 (2023-06-16 ucrt)"
```

## Importing Data 
```{r importing data}
hiv_data <- read_csv("hiv_trt.csv")

head(hiv_data)

```


## Missing Values Investigation 
```{r missing values}

#Taking care of missing variables 
summary(is.na(hiv_data))

# There are no missing values, if there were I would use some imputation method to account for them, make sure that they are not missign from primary outcome or treatment (or LOCF)

```

## Distribution of Categorical Variables
```{r dist of cat. variables}

# Summary table for Sex
sex_distribution <- hiv_data %>%
  count(sex)

# Summary table for Race/Ethnicity
race_eth_distribution <- hiv_data %>%
  count(raceth)

# Summary table for CD4 Stratum at Screening
strat2_distribution <- hiv_data %>%
  count(strat2)

# Summary table for IV drug use history
iv_drug_distribution <- hiv_data %>%
  count(ivdrug)

# Summary table for Hemophiliac
hemophil_distribution <- hiv_data %>%
  count(hemophil)

# Summary table for Karnofsky Performance Scale
karnof_scale_distribution <- hiv_data %>%
  count(karnof)

# Summary table for Treatment Groups
txgrp_distribution <- hiv_data %>%
  count(txgrp)

# Summary table for Treatment Indicator
tx_distribution <- hiv_data %>%
  count(tx)


print(sex_distribution)
print(race_eth_distribution)
print(strat2_distribution)
print(iv_drug_distribution)
print(hemophil_distribution)
print(karnof_scale_distribution)
print(txgrp_distribution)
print(tx_distribution)



```

## Recategorization 
```{r recategorizing}

#Combining variables to new categories for the ones that have low categories 
hiv_data <- hiv_data %>%
  mutate(race_eth = ifelse(raceth %in% c("3", "4", "5", "6"), "3", raceth),
         iv_drug = ifelse(ivdrug %in% c("2", "3"), "2", ivdrug),
         karnof_scale = ifelse(karnof %in% c("70", "80"), "80", karnof))
# Checking if Recategorization done correctly:
#Race

# New Race Count 
race_counts <- hiv_data %>%
  group_by(race_eth) %>%
  summarize(count = n())

# Print the results
print(race_counts)

# Old Race Count 
raceth_counts <- hiv_data %>%
  group_by(raceth) %>%
  summarize(count = n())

# Print the results
print(raceth_counts)

#IV Drug Use

# New IV Drug Use 
iv_drug_counts <- hiv_data %>%
  group_by(iv_drug) %>%
  summarize(count = n())

# Print the results
print(iv_drug_counts)

# Old IV Drug Use 
ivdrug_counts <- hiv_data %>%
  group_by(ivdrug) %>%
  summarize(count = n())

# Print the results
print(ivdrug_counts)

#Karnof Scale

# New Karnof
karnof_scale_counts <- hiv_data %>%
  group_by(karnof_scale) %>%
  summarize(count = n())

# Print the results
print(karnof_scale_counts)

# Old Karnof
karnof_counts <- hiv_data %>%
  group_by(karnof) %>%
  summarize(count = n())

# Print the results
print(karnof_counts)

# Deleting varaibles for old categories 
hiv_data <- hiv_data %>%
  select(-raceth, -ivdrug, -karnof)

```


## Labeling Categories
```{r labeling categories}

#Sex
hiv_data <- hiv_data %>%
  mutate(sex = case_when(
    sex == 1 ~ "Male",
    sex == 2 ~ "Female"))

#Race/Ethnicity
hiv_data <- hiv_data %>%
  mutate(race_eth = case_when(
    race_eth == 1 ~ "White Non-Hispanic",
    race_eth == 2 ~ "Black Non-Hispanic",
    race_eth == 3 ~ "Other"))

#CD4 Stratum at Screening
hiv_data <- hiv_data %>%
  mutate(strat2 = case_when(
    strat2 == 0 ~ "CD4 ≤ 50",
    strat2 == 1 ~ "CD4 > 50"))

#IV drug use history 
hiv_data <- hiv_data %>%
  mutate(iv_drug = case_when(
    iv_drug == 1 ~ "Never",
    iv_drug == 2 ~ "Currently/Previously"))

#Hemophiliac
hiv_data <- hiv_data %>%
  mutate(hemophil = case_when(
    hemophil == 0 ~ "No",
    hemophil == 1 ~ "Yes"))

#Karnofsky Performance Scale
hiv_data <- hiv_data %>%
  mutate(karnof_scale = case_when(
    karnof_scale == "80" ~ "70 or 80",
    karnof_scale == "90" ~ "90", 
    karnof_scale == "100" ~ "100"))

#Factoring Karnofsky Performance Scale
hiv_data$karnof_scale <- factor(hiv_data$karnof_scale,
                                 levels = c("70 or 80", "90", "100"),
                                 labels = c("70 or 80", "90", "100"))
#Treatment Groups
hiv_data <- hiv_data %>%
  mutate(txgrp = case_when(
    txgrp == 1 ~ "ZDV+3TC",
    txgrp == 2 ~ "ZDV+3TC+IDV", 
    txgrp == 3 ~ "d4T+3TC", 
    txgrp == 4 ~ "d4T+3TC+IDV"))

#Treatment Indicator
hiv_data <- hiv_data %>%
  mutate(tx= case_when(
    tx == 0 ~ "Control Group",
    tx == 1 ~ "Treatment includes IDV"))

```


## Distribution of Continuous Variables
```{r distributions continuous}
# Age - Normal - display mean
ggplot(data = hiv_data, aes(x = age)) +
  geom_histogram(binwidth = 5, fill = "skyblue", color = "black") +
  labs(title = "Histogram of Age",
       x = "Age",
       y = "Frequency")

# cd4 - right skewed - display median (mix/max or q1/q3?)
ggplot(data = hiv_data, aes(x = cd4)) +
  geom_histogram(binwidth = 5, fill = "skyblue", color = "black") +
  labs(title = "Histogram of cd4",
       x = "cd4",
       y = "Frequency")

# priorzdv - right skewed - display median (mix/max or q1/q3?)
ggplot(data = hiv_data, aes(x = priorzdv)) +
  geom_histogram(binwidth = 5, fill = "skyblue", color = "black") +
  labs(title = "Histogram of priorzdv",
       x = "priorzdv",
       y = "Frequency")

```


## Distribution of Categorical Variables (post-recategorization)
``` {r distributions categorical}

# Investigating categorical variables 
summary_stats_categorical <- function(cat_variable, t) {
  # Frequency Table
  freq_table <- table(cat_variable)
  print(freq_table)
  
  # Bar Plot
  barplot(freq_table, 
          main = t,
          xlab = "Category",
          ylab = "Frequency")
}


# Call the function with your binary variable
summary_stats_categorical(hiv_data$sex, "Frequency Distribution of Gender")
summary_stats_categorical(hiv_data$race_eth, "Frequency Distribution of Race")
summary_stats_categorical(hiv_data$strat2, " Frequency Distribution of  CD4 stratum at screening")
summary_stats_categorical(hiv_data$iv_drug, " Frequency Distribution of IV drug use history")
summary_stats_categorical(hiv_data$hemophil, "Frequency Distribution of  Hemophilia Status")
summary_stats_categorical(hiv_data$karnof_scale, "Frequency Distribution of Karnofsky Performance Scale")
summary_stats_categorical(hiv_data$txgrp, "Frequency Distribution of  Treatment group indicator")
summary_stats_categorical(hiv_data$tx, "Frequency Distribution of Treatment indicator")

```


## Table 1
```{r table 1}

# Reproducing Table using gtsummary:-- not including txgrp, only has 3 obsevations

descriptive_table <- hiv_data %>%
  select(age, cd4, priorzdv, sex, race_eth, strat2, iv_drug, hemophil, karnof_scale, tx) %>%
  tbl_summary(
    by = tx, 
    statistic = list(age ~ "{mean} ({sd})",
                     cd4 ~ "[{min}, {median}, {max}]",
                     priorzdv ~ "[{min}, {median}, {max}]"),
    type = list(hemophil ~ "categorical", 
                strat2 ~ "categorical"),
    label = list(
      age ~ "Age",
      sex ~ "Sex",
      race_eth ~ "Race/Ethnicity",
      strat2 ~ "CD4 Stratum at Screening",
      iv_drug ~ "IV drug use history",
      hemophil ~ "Hemophiliac", 
      karnof_scale ~ "Karnofsky Performance Scale", 
      cd4 ~ "Baseline CD4 count 
(Cells/milliliter)",
priorzdv ~ "Months of prior ZDV use"
)
  ) %>%
  add_overall() %>%
  add_p()

# Print the summary table
descriptive_table

# Remove tx from predictors 

```


## PFS: Log Rank Test and KM Curve
``` {r PFS log rank and KM}

# PFS
# log-rank test -- is there significant difference in survival curves between treatment groups? p= 0.00117 so yes, there is 

print(survdiff(Surv(time, censor) ~ tx, data = hiv_data),
      digits=5)


# Calculate the survival rates by treatment groups 
survival_by_tx <- survfit(Surv(time, censor) ~ tx,
                            data = hiv_data,
                            type = "kaplan-meier", 
                            conf.type = "log-log") 
print(survival_by_tx)
summary(survival_by_tx)

#Plotting K-M curves
survfit2(Surv(time, censor) ~ tx, data = hiv_data) %>%
  ggsurvfit() +
  labs (
    x = "Days",
    y = "Progression Free survival probability",
  ) +
  add_pvalue(location="annotation",
             caption="Log-rank {p.value}")


```

## PFS: Univariate Analysis
``` {r Cox univariate analysis PFS}

# Cox Proportinal Hazard Model to assess the association between predictor of interest and survival
#Univariate Analysis:

fitModel1 = coxph(Surv(time, censor) ~ tx, data = hiv_data,ties="efron")
summary(fitModel1)
tbl_regression(fitModel1, exponentiate = TRUE)


fitModel1grp = coxph(Surv(time, censor) ~ txgrp, data = hiv_data,ties="efron")
summary(fitModel1grp)
tbl_regression(fitModel1grp, exponentiate = TRUE)

fitModel1sex = coxph(Surv(time, censor) ~ sex, data = hiv_data,ties="efron")
summary(fitModel1sex)
tbl_regression(fitModel1sex, exponentiate = TRUE)


fitModel1age = coxph(Surv(time, censor) ~ age, data = hiv_data,ties="efron")
summary(fitModel1age)
tbl_regression(fitModel1age, exponentiate = TRUE)


fitModel1raceth = coxph(Surv(time, censor) ~ race_eth, data = hiv_data,ties="efron")
summary(fitModel1raceth)
tbl_regression(fitModel1raceth, exponentiate = TRUE)


fitModel1ivdrug = coxph(Surv(time, censor) ~ iv_drug, data = hiv_data,ties="efron")
summary(fitModel1ivdrug)
tbl_regression(fitModel1ivdrug, exponentiate = TRUE)


fitModel1hemophil = coxph(Surv(time, censor) ~ hemophil, data = hiv_data,ties="efron")
summary(fitModel1hemophil)
tbl_regression(fitModel1hemophil, exponentiate = TRUE)


fitModel1karnof = coxph(Surv(time, censor) ~ karnof_scale, data = hiv_data,ties="efron")
summary(fitModel1karnof)
tbl_regression(fitModel1karnof, exponentiate = TRUE)


fitModel1strat2 = coxph(Surv(time, censor) ~ strat2, data = hiv_data,ties="efron")
summary(fitModel1strat2)
tbl_regression(fitModel1strat2, exponentiate = TRUE)


fitModel1cd4 = coxph(Surv(time, censor) ~ cd4, data = hiv_data,ties="efron")
summary(fitModel1cd4)
tbl_regression(fitModel1cd4, exponentiate = TRUE)


fitModel1priorzdv = coxph(Surv(time, censor) ~ priorzdv, data = hiv_data,ties="efron")
summary(fitModel1priorzdv)
tbl_regression(fitModel1priorzdv, exponentiate = TRUE)

```

## PFS: Multivariate Analysis
```{r multivariate analysis PFS}

#Multivariate Analysis 
#Fitting cox model for all variables to assess the association between predictor of interest and survival adjusting for other variables 
fitcoxall = coxph(Surv(time, censor) ~ tx + sex + age + race_eth + iv_drug + hemophil + karnof_scale +
                                      strat2 + cd4 + priorzdv, data = hiv_data,ties = "efron")
summary(fitcoxall)
tbl_regression(fitcoxall, exponentiate = TRUE)


#Fitting cox model for all significant variables in univariate analysis (karnof_scale, start2, cd4) + sex and age

fitcox = coxph(Surv(time, censor) ~ tx + sex + age + karnof_scale + strat2 + cd4,
                                    data = hiv_data,ties = "efron")

summary(fitcox)
tbl_regression(fitcox, exponentiate = TRUE)

```

## PFS: Model Selection
``` {r model selection PFS}

# compare 2 models -- fitModel1 vs fitcox
# LRT between univariate (fitModel1) vs model with significant + sex + age (fitcox)

#LRT Model 1 - univariate with just treatment:
fitModel1 = coxph(Surv(time, censor)~tx, data = hiv_data,ties="efron")
tbl_regression(fitModel1, exponentiate = TRUE)

# Calculating the log-likelihood of fitModel1 to see how well the model fits the data.
logLik(fitModel1)

#LRT Model - all significant variables in univariate analysis (karnof_scale, start2, cd4) + sex and age
fitcox = coxph(Surv(time, censor) ~ tx + sex + age + karnof_scale + strat2 + cd4, 
                                    data = hiv_data,ties = "efron")
# Calculating the log-likelihood of fitcox to see how well the model fits the data
logLik(fitcox)

#Comparing both models:
1-pchisq(as.numeric(-2*(logLik(fitModel1)-logLik(fitcox))),df=5)

#The degrees of freedom for the LRT is the difference in the number of parameters between the two models. (6-1 = 5)
# P-value is below 0.05, so rejecting the null hypothesis in favor of the alternative, suggesting that the more complex model (fitcox) provides a significantly better fit than the simpler model (fitModel1).

# compare 2 models -- fitcoxfinal vs fitcox

# Final Model Selection prior to interaction testing --> (only include significant variables from multivariate above vs original multivatiate)
fitcoxfinal = coxph(Surv(time, censor) ~ tx + sex + age + karnof_scale + cd4,
                                        data = hiv_data,ties = "efron")
summary(fitcoxfinal)
tbl_regression(fitcoxfinal, exponentiate = TRUE)

# Calculating the log-likelihood of fitcoxfinal and fitcox to see how well the model fits the data
logLik(fitcoxfinal)
logLik(fitcox)

#Comparing both models:
1-pchisq(as.numeric(-2*(logLik(fitcox)-logLik(fitcoxfinal))),df=1)

# Result, no difference between them, so choosing more parsimonious model w/o start2 --> fitcoxfinal

```

## PFS: Interaction Tests 
```{r interaction testing PFS}
# Add interaction term for significant variables 

# Interaction with age
fitModelInt = coxph(Surv(time, censor) ~ tx + sex + age + karnof_scale + cd4 + tx*age,
                                        data = hiv_data, ties = "efron")

logLik(fitModelInt)
logLik(fitcoxfinal)
1-pchisq(as.numeric(-2*(logLik(fitcox)-logLik(fitModelInt))),df=1)

# Interaction with karnof
fitModelInt1 = coxph(Surv(time, censor) ~ tx + sex + age + karnof_scale + cd4 + tx*karnof_scale,
                                        data = hiv_data, ties = "efron")

logLik(fitModelInt1)
logLik(fitcoxfinal)
1-pchisq(as.numeric(-2*(logLik(fitcox)-logLik(fitModelInt1))),df=1)

# Interaction with cd4
fitModelInt2 = coxph(Surv(time, censor) ~ tx + sex + age + karnof_scale + cd4 + tx*cd4,
                                        data = hiv_data, ties = "efron")

logLik(fitModelInt2)
logLik(fitcoxfinal)
1-pchisq(as.numeric(-2*(logLik(fitcox)-logLik(fitModelInt2))),df=1)

# Interaction with sex
fitModelInt3 = coxph(Surv(time, censor) ~ tx + sex + age + karnof_scale + cd4 + tx*sex,
                                        data = hiv_data, ties = "efron")

logLik(fitModelInt3)
logLik(fitcoxfinal)
1-pchisq(as.numeric(-2*(logLik(fitcox)-logLik(fitModelInt3))),df=1)
#tested all interactions and there is no significant difference in models 

```

## PFS: Final Model Selected + Results
```{r final model PFS}

# Final Model selected: 
fitcoxfinal <-  coxph(Surv(time, censor) ~ tx + sex + age + karnof_scale + cd4,
                                        data = hiv_data, ties = "efron")
summary(fitcoxfinal)

tbl_regression(fitcoxfinal, exponentiate = TRUE)

```

## OS: Log Rank Test and KM Curve
```{r overall survival}

# Overall Survival 

# log-rank test
print(survdiff(Surv(time_d, censor_d) ~ tx, data = hiv_data),
      digits=5)

# Calculate the survival rates by treatment groups 
survival_by_tx_OS <- survfit(Surv(time_d, censor_d) ~ tx,
                            data = hiv_data,
                            type = "kaplan-meier", 
                            conf.type = "log-log") 
print(survival_by_tx_OS)
summary(survival_by_tx_OS)


#Plotting K-M curves
survfit2(Surv(time_d, censor_d) ~ tx, data = hiv_data) %>%
  ggsurvfit() +
  labs (
    x = "Days",
    y = "Overall survival probability",
  ) +
  add_pvalue(location="annotation",
             caption="Log-rank {p.value}")

```

## OS: Univariate Analysis
```{r univariate analysis OS}

#Univariate Analysis:
fitModel1OS = coxph(Surv(time_d, censor_d)~tx, data=hiv_data,ties="efron")
summary(fitModel1OS)
tbl_regression(fitModel1OS, exponentiate = TRUE)

fitModel1OStxgrp = coxph(Surv(time_d, censor_d)~txgrp, data=hiv_data,ties="efron")
summary(fitModel1OStxgrp)
tbl_regression(fitModel1OStxgrp, exponentiate = TRUE)

fitModel1sexOS = coxph(Surv(time_d, censor_d)~ sex, data=hiv_data,ties="efron")
summary(fitModel1sexOS)
tbl_regression(fitModel1sexOS, exponentiate = TRUE)

fitModel1ageOS = coxph(Surv(time_d, censor_d)~ age, data=hiv_data,ties="efron")
summary(fitModel1ageOS)
tbl_regression(fitModel1ageOS, exponentiate = TRUE)

fitModel1racethOS = coxph(Surv(time_d, censor_d)~ race_eth, data=hiv_data,ties="efron")
summary(fitModel1racethOS)
tbl_regression(fitModel1racethOS, exponentiate = TRUE)

fitModel1ivdrugOS = coxph(Surv(time_d, censor_d)~ iv_drug, data=hiv_data,ties="efron")
summary(fitModel1ivdrugOS)
tbl_regression(fitModel1ivdrugOS, exponentiate = TRUE)

fitModel1hemophilOS = coxph(Surv(time_d, censor_d)~ hemophil, data=hiv_data,ties="efron")
summary(fitModel1hemophilOS)
tbl_regression(fitModel1hemophilOS, exponentiate = TRUE)

fitModel1karnofOS = coxph(Surv(time_d, censor_d)~ karnof_scale, data=hiv_data,ties="efron")
summary(fitModel1karnofOS)
tbl_regression(fitModel1karnofOS, exponentiate = TRUE)

fitModel1strat2OS = coxph(Surv(time_d, censor_d)~ strat2, data=hiv_data,ties="efron")
summary(fitModel1strat2OS)
tbl_regression(fitModel1strat2OS, exponentiate = TRUE)

fitModel1cd4OS = coxph(Surv(time_d, censor_d)~ cd4, data=hiv_data,ties="efron")
summary(fitModel1cd4OS)
tbl_regression(fitModel1cd4OS, exponentiate = TRUE)

fitModel1priorzdvOS = coxph(Surv(time_d, censor_d)~ priorzdv, data=hiv_data,ties="efron")
summary(fitModel1priorzdvOS)
tbl_regression(fitModel1priorzdvOS, exponentiate = TRUE)

```

## OS: Multivariate Analysis
```{r multivariate analysis OS}

# Multivariate Analysis w/ all covariates 
fitcoxOSall = coxph(Surv(time_d, censor_d) ~ tx + sex + age + race_eth + iv_drug + hemophil + karnof_scale + strat2 + cd4 + priorzdv,
               data=hiv_data,ties="efron")
tbl_regression(fitcoxOSall, exponentiate = TRUE)

# Only Significant + sex and age
fitcoxOS = coxph(Surv(time_d, censor_d) ~ tx + sex + age + karnof_scale + strat2 + cd4,
                 data=hiv_data,ties="efron")
summary(fitcoxOS)
tbl_regression(fitcoxOS, exponentiate = TRUE)


```

## OS: Model Selection
```{r model selection OS}

# compare 2 models --> fitModel1OS vs fitcoxOS
# LRT 
#LRT Model 1:
fitModel1OS = coxph(Surv(time_d, censor_d) ~ tx, data = hiv_data,ties="efron")

#LRT Univariate 
logLik(fitModel1OS)

# LRT Only Significant + sex and age
logLik(fitcoxOS)

#Compare both
1-pchisq(as.numeric(-2*(logLik(fitModel1OS)-logLik(fitcoxOS))),df=5)


# Final Model Selection prior to interaction testing --  (only include significant variables from multivariate above vs original multivatiate)

# Only significant from multivariate
fitcoxfinalOS = coxph(Surv(time, censor) ~ tx + sex + age + karnof_scale,
                                          data = hiv_data, ties = "efron")
summary(fitcoxfinalOS)
tbl_regression(fitcoxfinalOS, exponentiate = TRUE)


# LRT for each model 
logLik(fitcoxfinalOS)
logLik(fitcoxOS)

#Compare them 
1-pchisq(as.numeric(-2*(logLik(fitcoxOS)-logLik(fitcoxfinalOS))),df=2)
# Result, no difference between them, so choosing more parsimonious

```

## OS: Interaction Tests 
```{r OS interaction testing}
# Add interaction term for significant variables 
# Interaction with age
fitModelIntOS = coxph(Surv(time_d, censor_d) ~ tx + sex + age + karnof_scale + tx*age,
                                              data = hiv_data, ties = "efron")

logLik(fitModelIntOS)
logLik(fitcoxOS)
1-pchisq(as.numeric(-2*(logLik(fitcoxOS)-logLik(fitModelIntOS))),df=1)

# Interaction with karnof
fitModelInt1OS = coxph(Surv(time_d, censor_d) ~ tx + sex + age + karnof_scale + tx*karnof_scale,
                                              data = hiv_data, ties = "efron")

logLik(fitModelInt1OS)
logLik(fitcoxOS)
1-pchisq(as.numeric(-2*(logLik(fitcoxOS)-logLik(fitModelInt1OS))),df=1)

# Interaction with sex
fitModelInt2OS = coxph(Surv(time_d, censor_d) ~ tx + sex + age + karnof_scale + tx*sex,
                                            data = hiv_data, ties="efron")

logLik(fitModelInt2OS)
logLik(fitcoxOS)
1-pchisq(as.numeric(-2*(logLik(fitcoxOS)-logLik(fitModelInt2OS))),df=1)

#tested all interactions and there is no significant difference between models 

```

## OS: Final Model Selected + Results
```{r final OS model}
# Final Model selected
fitcoxfinalOS = coxph(Surv(time_d, censor_d) ~ tx + sex + age + karnof_scale,
                                            data = hiv_data, ties = "efron")
summary(fitcoxfinalOS)

tbl_regression(fitcoxfinalOS, exponentiate = TRUE)

```
